A Dependently Typed Multi-Stage Calculus

### 2019-08-29 text/formal-defnition.tex
Towards the camera ready version.

The prefix operators $\TW_\alpha, \TB_\alpha, \TBL_\alpha$, and
$\%_\alpha$ connect tighter than the three forms $\tau\ M$, $M\ N$,
$M\ A$ of applications, which are left-associative. The binders $\Pi$,
$\forall$, and $\lambda$ extend as far to the right as possible.
Thus, $\F\alpha.\TW_{\alpha} (\Pi x:\I.\text{Vector}\ 5)$ is
interpreted as
$\F\alpha.(\TW_{\alpha} (\Pi x:\I.(\text{Vector}\ 5)))$; and
$\Lambda\alpha.\lambda x:\I.\TB_\alpha x\ y$ means
$\Lambda\alpha.(\lambda x:\I.(\TB_\alpha x)\ y)$.

\paragraph{Remark:} Basically, we define \LMD to be an extension of
\LTP with dependent types.  One notable difference is that \LMD has
only one kind of \(\alpha\)-closed types, whereas \LTP has two kinds
of \(\alpha\)-closed types \(\forall\alpha.\tau\) and
\(\forall^\varepsilon\alpha.\tau\).  We have omitted the first kind,
for simplicity, and dropped the superscript $\varepsilon$ from the second. It
would not be difficult to recover the distinction to show properties related
to program residualization~\cite{Hanada2014}, although they are still
conjecture.

@@

Next, we define full reduction for \LMD.
Before giving the definition of reduction, we define two kinds of substitutions.
Substitution $M[x\mapsto N], \tau[x \mapsto N]$ and $K[x \mapsto N]$ are
ordinary capture-avoiding substitution of
term $N$ for $x$ in term $M$, type $\tau$, and kind $K$, respectively,
and we omit their definitions here.
Substitution $M[\alpha \mapsto A], \tau [\alpha \mapsto A], K[\alpha \mapsto A]$ and $B[\alpha\mapsto A]$ are
substitutions of stage $A$ for stage variable $\alpha$ in
term $N$ for $x$ in term $M$, type $\tau$, kind $K$, and stage $B$, respectively.
We show representative cases below.
%
{%\small
\begin{align*}
    (\lambda x:\tau.M)[\alpha \mapsto A] & = \lambda x:(\tau[\alpha \mapsto A]).(M[\alpha \mapsto A])                                  \\
    (M\ B)[\alpha \mapsto A]             & = (M[\alpha \mapsto A])\ B[\alpha\mapsto A]                                                 \\
    (\TB_\beta M)[\alpha \mapsto A]      & = \TB_{\beta[\alpha \mapsto A]}M[\alpha \mapsto A]                                          \\
    (\TBL_\beta M)[\alpha \mapsto A]     & = \TBL_{\beta[\alpha \mapsto A]}M[\alpha \mapsto A]                                         \\
    (\%_\beta M)[\alpha \mapsto A]       & = \%_{\beta[\alpha \mapsto A]}M[\alpha \mapsto A]                                           \\
    (\beta B)[\alpha \mapsto A]          & = \beta (B[\alpha\mapsto A])                               & (\text{if } \alpha \neq \beta) \\
    (\beta B)[\alpha \mapsto A]          & = A (B[\alpha\mapsto A])                                   & (\text{if } \alpha = \beta)
\end{align*}
}
Here, $\TB_{\alpha_1\cdots\alpha_n} M$,
$\TBL_{\alpha_1\cdots\alpha_n} M$, and $\%_{\alpha_1\cdots\alpha_n} M$
$(n \geq 0)$ stand for $\TB_{\alpha_1} \cdots \TB_{\alpha_n} M$,
$\TBL_{\alpha_1}\cdots \TBL_{\alpha_n} M$, and
$\%_{\alpha_1}\cdots \%_{\alpha_n} M$, respectively.  In particular,
$\TB_{\varepsilon} M = \TBL_{\varepsilon} M = \%_{\varepsilon} M = M$.
We also define substitutions of a stage or a term for variables in type environment $\G$.

@@

The rules for kind well-formedness and kinding is a straightforward
adaptation from \LLF and \LTP, except for the following rule for type-level CSP.
\begin{center}
  \infrule[K-Csp]{
    \G \V \tau :: * @ A
  }{
    \G \V \tau :: * @ A\alpha
  }
\end{center}
Unlike the term level, type-level CSP is implicit because there is no staged
semantics for types. \AK{This reason may not be enough for justification.}

@@

% Typing rules for a multi-stage calculus
The rules \TTB, \TTBL, \TGen, \TIns, and \TCsp{} are constructs for
multi-stage progarmming. \TTB{} and \TTBL{} are the same as \LTP, as we
explained in Section \ref{sec:informal-overview}. The rule \TGen{} for stage
abstraction is straightforward. The condition
$\alpha\notin\rm{FTV}(\G)\cup\rm{FTV}(A)$ ensures that the scope of $\alpha$
is in $M$, and avoids capturing variables in the rest of the judgments. The
rule \TIns{} is for applications of stages to stage abstractions. The rule
\TCsp{} is for CSP, which means that, if term $M$ is of type $\tau$ at stage
$A$, then $\%_\alpha M$ is of type $\tau$ at stage $A\alpha$. Note that CSP
is also applied to the type \(\tau\) (although it is implicit) in the
conclusion. Thanks to implicit CSP, the typing rules are the same as in \LTP.

@@

This rule is motivated by a practical consideration. We have potentially many
pairs of vectors of the fixed length in some fields, for example, computer
graphics and the same function such as vector addition is applied to all of
them. \AK{I strenghened this example by adding concrete situation. Do we need
more reinforcements such as comparision with other method?} We will discuss
the motivation through an example of specializing a vector addition function
with respect to the size of vectors. The function vadd (with an auxiliary
function vadd$_1$) below implements a code generator in a hypothetical
language based on \LTP. This language provides definitions by \textbf{let},
recursive functions (represented by \textbf{fix}), \textbf{if}-expressions,
and primitives cons, head, and tail to manipulate vectors. We assume that
$\text{cons}$ is of type $\Pi n:\I.\I \to \text{Vector}\ n \to \text{Vector}\ (n+1)$, 
$\text{head}$ is of type $\Pi n:\I.\text{Vector}\ (n+1) \to \I$, and
$\text{tail}$ is of type $\Pi n:\I.\text{Vector}\ (n+1) \to (\text{Vector}\ n)$.

@@

The main function vadd below takes the length $n$ and returns
($\beta$-closed) code of a function that takes two vectors and returns
their addition using vadd$_1$.
\renewcommand{\Vpn}{\text{Vector}\ (\%_\beta n)}
\begin{align*}
	  & \textbf{let}\ \text{vadd}: \Pi n:\I.\F\beta.\TW_\beta(\Vpn\to\Vpn\to\Vpn)                \\ 
	  & \hspace{6mm} = \lambda n:\I.\Lambda\beta.\TB_\beta (\lambda v_1:\Vpn.\ \lambda v_2:\Vpn. \\
	  & \hspace{63mm} \TBL_\beta (\text{vadd}_1\ \beta\ n\ (\TB_\beta\ v_1)\ (\TB_\beta\ v_2))) 
\end{align*}
\renewcommand{\Vpn}{\text{Vector}\ (\%_\beta 5)}%
If this function is applied to a
(nonnegative) integer constant, say 5, it returns function code for adding
two vectors of size 5.  The type of vadd\ 5, obtained by
substituting 5 for $n$, is
$\F\beta.\TW_\beta(\Vpn\to\Vpn\to\Vpn)$.
\renewcommand{\Vpn}{\text{Vector}\ 5}
If the obtained code is run by applying to \(\varepsilon\),
the type of vadd\ 5\ $\varepsilon$ is
\(\Vpn\to\Vpn\to\Vpn\) as expected.
Without \QPercent, however, it would not be possible
to compose this function code with another piece of code
$v$ of type, say, \(\TW_\gamma \text{Vector }5\) because
\(\text{Vector }5\) and \(\text{Vector }(\%_\beta 5)\) are
syntactically different types.
% : \(\TB_\gamma (\TBL_\gamma (\text{vadd}\ 5\ \gamma)\ v\ v)\).

Interestingly, Hanada and Igarashi~\cite{Hanada2014} rejected the idea of
reduction that removes $\%_\alpha$ when they developed \LTP{}, as such
reduction does not match the operational behavior of the CSP operator
in implementations.  However, as an equational system for multi-stage
programs, the rule \QPercent makes sense.

@@

\begin{definition}[Evaluation Context]
  Let $B$ be either \(\varepsilon\) or a stage variable \(\beta\).
  The family of sets $ECtx^A_B$ of evaluation contexts, ranged over by $E^A_B$, is defined by the following grammar (in which $A'$ stands for a non-empty stage).
  %  $A$ is assumed to be nonempty (but $B,A'$ can be empty).
  \begin{align*}
    E^\varepsilon_B \in ECtx^\varepsilon_B & ::= \square\ (\text{if\ } B = \varepsilon)
    \mid E^\varepsilon_B\ M \mid v^\varepsilon\ E^\varepsilon_B \mid \TB_\alpha E^\alpha_B
    \mid \Lambda\alpha.E^\varepsilon_B \mid E^\varepsilon_B\ A                                                                                    \\
    E^{A'}_B \in ECtx^{A'}_B               & ::= \square\ (\text{if } A' = B) \mid \lambda x:\tau.E^{A'}_B \mid E^{A'}_B\ M \mid v^{A'}\ E^{A'}_B \\
                                           & \mid \TB_\alpha E^{A'\alpha}_B \mid \TBL_\alpha E^{A}_B \ (\text{where } A\alpha = A')               \\
                                           & \mid \Lambda\alpha.E^{A'}_B \mid E^{A'}_B\ A \mid \%_\alpha\ E^{A'}_B \ (\text{where } A\alpha = A')
  \end{align*}
\end{definition}

The subscripts $A$ and $B$ in $E^A_B$ stand for the stage of the evaluation
context and of the hole, respectively. The grammar represents that staged
reduction is left-to-right and call-by-value and terms under \(\Lambda\) are
reduced. Terms at non-$\varepsilon$ stages are not reduced as long as they do
not contain a term such as \(\TBL_\alpha \TB_\alpha M\). A few examples of
evaluation context are shown below:
\begin{align*}
  \square\ (\lambda x:\I.x)                  & \in  ECtx^\varepsilon_\varepsilon \\
  \Lambda\alpha.\square\ \epsilon            & \in ECtx^\varepsilon_\varepsilon  \\
  \TBL_\alpha \TB_\alpha \TBL_\alpha \square & \in ECtx^\alpha_\varepsilon
\end{align*}
%
We write $E^A_B[M]$ for the term obtained by filling the hole $\square$ in $E^A_B$ by $M$.

@@

This reduction relation represents a term reduces in a deterministic,
left-to-right, call-by-value manner.  An application of an abstraction
is executed only at stage \(\varepsilon\) and only a quotation at
stage \(\varepsilon\) is spliced into the surrounding code---notice
that, if \(\TB_\alpha v^\alpha\) is at stage \(\varepsilon\), then the
redex \(\TBL_\alpha \TB_\alpha v^\alpha\) is at stage \(\alpha\).
In other words, terms in brackets are not evaluated until the terms are run
and arguments of a function are evaluated before the application.
We show an example of staged reduction.
Underlines show the redexes.
\begin{align*}
                    & (\Lambda\alpha.(\TB_\alpha \underline{\TBL_\alpha \TB_\alpha ((\lambda x:\I.x)\ 10))})\ \varepsilon \\
  \longrightarrow_s & \underline{(\Lambda\alpha.(\TB_\alpha ((\lambda x:\I.x)\ 10)))\ \varepsilon}                        \\
  \longrightarrow_s & \underline{(\lambda x:\I.x)\ 10}                                                                    \\
  \longrightarrow_s & 10
\end{align*}

>>>>

The prefix operators $\TW_\alpha, \TB_\alpha, \TBL_\alpha$, and
$\%_\alpha$ are given higher precedence over the three forms $\tau\ M$, $M\ N$,
$M\ A$ of applications, which are left-associative. The binders $\Pi$,
$\forall$, and $\lambda$ extend as far to the right as possible.
Thus, $\F\alpha.\TW_{\alpha} (\Pi x:\I.\text{Vector}\ 5)$ is
interpreted as
$\F\alpha.(\TW_{\alpha} (\Pi x:\I.(\text{Vector}\ 5)))$; and
$\Lambda\alpha.\lambda x:\I.\TB_\alpha x\ y$ means
$\Lambda\alpha.(\lambda x:\I.(\TB_\alpha x)\ y)$.

\paragraph{Remark:} Basically, we define \LMD to be an extension of
\LTP with dependent types.  One notable difference is that \LMD has
only one kind of \(\alpha\)-closed types, whereas \LTP has two kinds
of \(\alpha\)-closed types \(\forall\alpha.\tau\) and
\(\forall^\varepsilon\alpha.\tau\).  We have omitted the first kind,
for simplicity, and dropped the superscript $\varepsilon$ from the second. It
would not be difficult to recover the distinction to show properties related
to program residualization~\cite{Hanada2014}, although they are left as 
conjectures.

@@

Next, we define full reduction for \LMD.
Before giving the definition of reduction, we define two kinds of substitutions.
Substitution $M[x\mapsto N], \tau[x \mapsto N]$ and $K[x \mapsto N]$ are
ordinary capture-avoiding substitution of
term $N$ for $x$ in term $M$, type $\tau$, and kind $K$, respectively,
and we omit their definitions here.
Substitution $M[\alpha \mapsto A], \tau [\alpha \mapsto A], K[\alpha \mapsto A]$ and $B[\alpha\mapsto A]$ are
substitutions of stage $A$ for stage variable $\alpha$ in term $M$, type $\tau$, kind $K$, and stage $B$, respectively.
We show representative cases below.
%
{%\small
\begin{align*}
    (\lambda x:\tau.M)[\alpha \mapsto A] & = \lambda x:(\tau[\alpha \mapsto A]).(M[\alpha \mapsto A])                                  \\
    (M\ B)[\alpha \mapsto A]             & = (M[\alpha \mapsto A])\ B[\alpha\mapsto A]                                                 \\
    (\TB_\beta M)[\alpha \mapsto A]      & = \TB_{\beta[\alpha \mapsto A]}M[\alpha \mapsto A]                                          \\
    (\TBL_\beta M)[\alpha \mapsto A]     & = \TBL_{\beta[\alpha \mapsto A]}M[\alpha \mapsto A]                                         \\
    (\%_\beta M)[\alpha \mapsto A]       & = \%_{\beta[\alpha \mapsto A]}M[\alpha \mapsto A]                                           \\
    (\beta B)[\alpha \mapsto A]          & = \beta (B[\alpha\mapsto A])                               & (\text{if } \alpha \neq \beta) \\
    (\beta B)[\alpha \mapsto A]          & = A (B[\alpha\mapsto A])                                   & (\text{if } \alpha = \beta)
\end{align*}
}
Here, $\TB_{\alpha_1\cdots\alpha_n} M$,
$\TBL_{\alpha_1\cdots\alpha_n} M$, and $\%_{\alpha_1\cdots\alpha_n} M$
$(n \geq 0)$ stand for $\TB_{\alpha_1} \cdots \TB_{\alpha_n} M$,
$\TBL_{\alpha_1}\cdots \TBL_{\alpha_n} M$, and
$\%_{\alpha_1}\cdots \%_{\alpha_n} M$, respectively.  In particular,
$\TB_{\varepsilon} M = \TBL_{\varepsilon} M = \%_{\varepsilon} M = M$.
We also define substitutions of a stage or a term for variables in type environment $\G$.

@@

The rules for kind well-formedness and kinding is a straightforward
adaptation from \LLF and \LTP, except for the following rule for type-level CSP.
\begin{center}
  \infrule[K-Csp]{
    \G \V \tau :: * @ A
  }{
    \G \V \tau :: * @ A\alpha
  }
\end{center}
Unlike the term level, type-level CSP is implicit because there is no staged
semantics for types.

@@

% Typing rules for a multi-stage calculus
The rules \TTB, \TTBL, \TGen, \TIns, and \TCsp{} are constructs for
multi-stage progarmming. \TTB{} and \TTBL{} are the same as in \LTP, as we
explained in Section \ref{sec:informal-overview}. The rule \TGen{} for stage
abstraction is straightforward. The condition
$\alpha\notin\rm{FTV}(\G)\cup\rm{FTV}(A)$ ensures that the scope of $\alpha$
is in $M$, and avoids capturing variables in the rest of the judgments. The
rule \TIns{} is for applications of stages to stage abstractions. The rule
\TCsp{} is for CSP, which means that, if term $M$ is of type $\tau$ at stage
$A$, then $\%_\alpha M$ is of type $\tau$ at stage $A\alpha$. Note that CSP
is also applied to the type \(\tau\) (although it is implicit) in the
conclusion. Thanks to implicit CSP, the typing rules are the same as in \LTP.

@@

\subsubsection{Example.}

We show an example of a dependently typed code generator in a
hypothetical language based on \LTP.  
% This rule is motivated by a practical consideration.
Let's consider an application, for example, in computer graphics, in
which we have potentially many pairs of vectors of the fixed (but
statically unkown) length and a function---such as vector
addition---to be applied to all of them.

\AI{Some more description about what we'd like to do.}
\AI{Brief comparison with other ways (somewhere else)?}
% \AK{I strenghened this example by adding concrete situation. Do we need
% more reinforcements such as comparision with other method?} We will discuss
% the motivation through an example of specializing a vector addition function
% with respect to the size of vectors.

The function vadd (with an auxiliary
function vadd$_1$) below implements a code generator.
This language provides definitions by \textbf{let},
recursive functions (represented by \textbf{fix}), \textbf{if}-expressions,
and primitives cons, head, and tail to manipulate vectors. We assume that
$\text{cons}$ is of type $\Pi n:\I.\I \to \text{Vector}\ n \to \text{Vector}\ (n+1)$, 
$\text{head}$ is of type $\Pi n:\I.\text{Vector}\ (n+1) \to \I$, and
$\text{tail}$ is of type $\Pi n:\I.\text{Vector}\ (n+1) \to (\text{Vector}\ n)$.

@@

The main function vadd below takes the length $n$ and returns
($\beta$-closed) code of a function that takes two vectors and returns
their addition using vadd$_1$.
\renewcommand{\Vpn}{\text{Vector}\ (\%_\beta n)}
\begin{align*}
	  & \textbf{let}\ \text{vadd}: \Pi n:\I.\F\beta.\TW_\beta(\Vpn\to\Vpn\to\Vpn)                \\ 
	  & \hspace{6mm} = \lambda n:\I.\Lambda\beta.\TB_\beta (\lambda v_1:\Vpn.\ \lambda v_2:\Vpn. \\
	  & \hspace{63mm} \TBL_\beta (\text{vadd}_1\ \beta\ n\ (\TB_\beta\ v_1)\ (\TB_\beta\ v_2))) 
\end{align*}
\renewcommand{\Vpn}{\text{Vector}\ (\%_\beta 5)}%
If this function is applied to a
(nonnegative) integer constant, say 5, it returns function code for adding
two vectors of size 5.  The type of vadd\ 5, obtained by
substituting 5 for $n$, is
$\F\beta.\TW_\beta(\Vpn\to\Vpn\to\Vpn)$.
\renewcommand{\Vpn}{\text{Vector}\ 5}
If the obtained code is run by applying to \(\varepsilon\),
the type of vadd\ 5\ $\varepsilon$ is
\(\Vpn\to\Vpn\to\Vpn\) as expected.

\AI{Comparison?}

\paragraph{Remark:}
\AI{Comments on \QPercent}
Without \QPercent, however, it would not be possible
to compose this function code with another piece of code
$v$ of type, say, \(\TW_\gamma \text{Vector }5\) because
\(\text{Vector }5\) and \(\text{Vector }(\%_\beta 5)\) are
syntactically different types.
% : \(\TB_\gamma (\TBL_\gamma (\text{vadd}\ 5\ \gamma)\ v\ v)\).
Interestingly, Hanada and Igarashi~\cite{Hanada2014} rejected the idea of
reduction that removes $\%_\alpha$ when they developed \LTP{}, as such
reduction does not match the operational behavior of the CSP operator
in implementations.  However, as an equational system for multi-stage
programs, the rule \QPercent makes sense.

@@

\begin{definition}[Evaluation Context]
  Let $B$ be either \(\varepsilon\) or a stage variable \(\beta\).
  The family of sets $ECtx^A_B$ of evaluation contexts, ranged over by $E^A_B$, is defined by the following grammar (in which $A'$ stands for a non-empty stage).
  %  $A$ is assumed to be nonempty (but $B,A'$ can be empty).
  \begin{align*}
    E^\varepsilon_B \in ECtx^\varepsilon_B & ::= \square\ (\text{if\ } B = \varepsilon)
    \mid E^\varepsilon_B\ M \mid v^\varepsilon\ E^\varepsilon_B \mid \TB_\alpha E^\alpha_B
    \mid \Lambda\alpha.E^\varepsilon_B \mid E^\varepsilon_B\ A                                                                                    \\
    E^{A'}_B \in ECtx^{A'}_B               & ::= \square\ (\text{if } A' = B) \mid \lambda x:\tau.E^{A'}_B \mid E^{A'}_B\ M \mid v^{A'}\ E^{A'}_B \\
                                           & \mid \TB_\alpha E^{A'\alpha}_B \mid \TBL_\alpha E^{A}_B \ (\text{where } A\alpha = A')               \\
                                           & \mid \Lambda\alpha.E^{A'}_B \mid E^{A'}_B\ A \mid \%_\alpha\ E^{A}_B \ (\text{where } A\alpha = A')
  \end{align*}
\end{definition}

The subscripts $A$ and $B$ in $E^A_B$ stand for the stage of the evaluation
context and of the hole, respectively. The grammar represents that staged
reduction is left-to-right and call-by-value and terms under \(\Lambda\) are
reduced. Terms at non-$\varepsilon$ stages are not reduced, except
redexes of the form \(\TBL_\alpha \TB_\alpha M\) at stage \(\alpha\).
A few examples of
evaluation contexts are shown below:
\begin{align*}
  \square\ (\lambda x:\I.x)                  & \in  ECtx^\varepsilon_\varepsilon \\
  \Lambda\alpha.\square\ \varepsilon            & \in ECtx^\varepsilon_\varepsilon  \\
  \TBL_\alpha \TB_\alpha \TBL_\alpha \square & \in ECtx^\alpha_\varepsilon
\end{align*}
%
We write $E^A_B[M]$ for the term obtained by filling the hole $\square$ in $E^A_B$ by $M$.

@@

This reduction relation reduces a term in a deterministic,
left-to-right, call-by-value manner.  An application of an abstraction
is executed only at stage \(\varepsilon\) and only a quotation at
stage \(\varepsilon\) is spliced into the surrounding code---notice
that, if \(\TB_\alpha v^\alpha\) is at stage \(\varepsilon\), then the
redex \(\TBL_\alpha \TB_\alpha v^\alpha\) is at stage \(\alpha\).
In other words, terms in brackets are not evaluated until the terms are run
and arguments of a function are evaluated before the application.
We show an example of staged reduction.
Underlines show the redexes.
\begin{align*}
                    & (\Lambda\alpha.(\TB_\alpha \underline{\TBL_\alpha \TB_\alpha ((\lambda x:\I.x)\ 10))})\ \varepsilon \\
  \longrightarrow_s & \underline{(\Lambda\alpha.(\TB_\alpha ((\lambda x:\I.x)\ 10)))\ \varepsilon}                        \\
  \longrightarrow_s & \underline{(\lambda x:\I.x)\ 10}                                                                    \\
  \longrightarrow_s & 10
\end{align*}

### 2019-08-29 text/informal-overview.tex
Towards the camera ready version.

At first, the restriction above sounds too severe, because a term like
\(\lambda x:\I. \TB_\alpha (\lambda y:\text{Vector }x.M) \), which models a
typical code generator which takes the size $x$ and returns code for vector
manipulation specialized to the given size, frequently appears. It seems crucial
for \(y\) to be given a type indexed by $x$. We can address this problem by
CSP---In fact, $\text{Vector }x$ is not well formed at $\alpha$ under
$x:\I@\varepsilon$ but $\text{Vector }(\%_\alpha x)$ is! So, we can still
write \(\lambda x:\I. \TB_\alpha (\lambda y:\text{Vector }(\%_\alpha x).M) \)
for the typical sort of code generators.

>>>>

At first, the restriction above sounds too severe, because a term like
\(\lambda x:\I. \TB_\alpha (\lambda y:\text{Vector }x.M) \), which models a
typical code generator which takes the size $x$ and returns code for vector
manipulation specialized to the given size, will be rejected. It seems crucial
for \(y\) to be given a type indexed by $x$. We can address this problem by
CSP---In fact, $\text{Vector }x$ is not well formed at $\alpha$ under
$x:\I@\varepsilon$ but $\text{Vector }(\%_\alpha x)$ is! So, we can still
write \(\lambda x:\I. \TB_\alpha (\lambda y:\text{Vector }(\%_\alpha x).M) \)
for the typical sort of code generators.

### 2019-08-29 text/introduction.tex
Towards the camera ready version.

\subsection{Our Work}
In this paper, we develop a new multi-stage calculus \LMD by extending
the existing multi-stage calculus \LTP\cite{Hanada2014} with dependent
types and study its properties.  We base our work on \LTP, in which
the four multi-stage constructs are handled slightly differently from
MetaOCaml, because its type system and semantics are arguably simpler
than \(\lambda^\alpha\)~\cite{taha2003environment}, which formalizes
the design of MetaOCaml more faithfully.  Dependent types are based on
\LLF~\cite{attapl}, which has one of the simplest forms of dependent
types.  Our technical contributions are summarized as follows:
\begin{itemize}
\item We give a formal definition of \LMD with its syntax, 
  type system and two kinds of reduction---full and staged reduction.
\item We show preservation, strong normalization, and confluence for
  full reduction; and show unique decomposition (and progress as its
  corollary) for staged reduction, which formalizes program execution.
\end{itemize}
The combination of multi-stage programming and dependent types has
been discussed by Brady and Hammond~\cite{brady2006dependently} but,
to our knowledge, our work is a first formal calculus of dependently
typed multi-stage programming.

>>>>

\subsection{Our Work}
In this paper, we develop a new multi-stage calculus \LMD by extending
the existing multi-stage calculus \LTP\cite{Hanada2014} with dependent
types and study its properties.  We base our work on \LTP, in which
the four multi-stage constructs are handled slightly differently from
MetaOCaml, because its type system and semantics are arguably simpler
than \(\lambda^\alpha\)~\cite{taha2003environment}, which formalizes
the design of MetaOCaml more faithfully.  Dependent types are based on
\LLF~\cite{attapl}, which has one of the simplest forms of dependent
types.  Our technical contributions are summarized as follows:
\begin{itemize}
\item We give a formal definition of \LMD with its syntax, type system and
two kinds of reduction: full reduction, allowing reduction of any redex,
including one under $\lambda$-abstraction and quotation, and staged reduction, a
small-step call-by-value operational semantics that is closer to the intended
multi-stage implementation.
\item We show preservation, strong normalization, and confluence for
  full reduction; and show unique decomposition (and progress as its
  corollary) for staged reduction, which formalizes program execution.
\end{itemize}
The combination of multi-stage programming and dependent types has
been discussed by Pasalic, Taha, and Sheard~\cite{pasalic2002tagless}
and Brady and Hammond~\cite{brady2006dependently} but, to our
knowledge, our work is a first formal calculus of dependently typed
multi-stage programming with all the key constructs~\cite{MetaML}.

### 2019-08-29 text/related-work.tex
Fix.
Towards the camera ready version.

MetaOCaml is a programming language with quoting, unquoting, run, and CSP.
Kiselyov~\cite{8384206} describes many applications of MetaOCaml, including
filtering in signal processing, matrix-vector product, and a DSL compiler.
Ur\cite{chlipala2010ur} is a programming language with metaprogramming, which
is focused on Web applications.

@@

Theoretical studies on multi-stage programming owe a lot to seminal work by
Davies and Pfenning~\cite{DaviesPfenning01JACM} and
Davies~\cite{davies1996temporal}, who found Curry-Howard correspondence
between multi-stage calculi and modal logic. In particular, Davies'
$\lambda^\circ$~\cite{davies1996temporal} has been a basis for several
multi-stage calculi with quasi-quotation. $\lambda^\circ$ did not have
operators for run and CSP; a few
studies~\cite{benaissa1999logical,MoggiTahaBenaissaSheard99ESOP} enhanced and
improved $\lambda^\circ$ towards the development of a type-safe multi-stage
calculus with quasi-quotation, run, and CSP, which were proposed by Taha and
Sheard as constructs for multi-stage programming~\cite{MetaML}. Chen and
Xi~\cite{chen2003meta} introduce a code type with types of free variables in
the code in order to prevent free variable evaluation. \AK{Is here
appropriate? Although this paper using the concept of dependent types, the
purpose is avoid open code problem. Therefore, I cite this paper here.}

Finally, Taha and Nielsen invented the concept of environment
classifiers~\cite{taha2003environment} and developed a typed calculus
$\lambda^\alpha$, which was equipped with all the features above in a type
sound manner and formed a basis of earlier versions of MetaOCaml. Different
approaches to type-safe multi-stage programming with slightly different
constructs for composing and running code values have been studied by Kim,
Yi, and Calcagno~\cite{DBLP:conf/popl/KimYC06} and Nanevski and
Pfenning~\cite{DBLP:journals/jfp/NanevskiP05}.

@@

Although there are studies on combinations of multi-stage programming and
other programming features such as mutable cells~\cite{kiselyov2016refined}
and control operators~\cite{KameyamaKiselyovShan09PEPM,oishi2017staging}, a
combination with dependent types has been little studied. \AK{little?} Brady
and Hammond~\cite{brady2006dependently} have discussed a combination of
dependently typed programming with staging in the style of MetaOCaml to
implement a staged interpreter, which is statically guaranteed to generate
well-typed code. However, they focused on concrete programming examples and
there is no theoretical investigation of the programming language they used.
Pasalic and Taha~\cite{pasalic2002tagless} designed \(\lambda_{H\circ}\) by
introducing the concept of stage into an existing dependent type system
\(\lambda_H\)~\cite{zhong2002certified}. However, there are not eval and CSP
in \(\lambda_{H\circ}\) because it was designed before the invention of
environment classifiers. In Forgarty, Pasalic, Siek and
Taha~\cite{fogarty2007concoqtion}, they extended the type system of MetaOCaml
with indexed types. With this extension, types can be indexed with a Coq term.

In order to make safer multi-stage programming language, 
Martin and Laurence~\cite{martin2015HGRTMP} gave program logic for
\(\text{Mini-ML}^\square_e\)~\cite{DaviesPfenning01JACM} but
it cannot manipulate open code which ours can deal with. \AK{Is the place of
this sentence is appropriate?}

>>>>

MetaOCaml is a programming language with quoting, unquoting, run, and CSP.
Kiselyov~\cite{8384206} describes many applications of MetaOCaml, including
filtering in signal processing, matrix-vector product, and a DSL compiler.

@@

Theoretical studies on multi-stage programming owe a lot to seminal work by
Davies and Pfenning~\cite{DaviesPfenning01JACM} and
Davies~\cite{davies1996temporal}, who found Curry-Howard correspondence
between multi-stage calculi and modal logic. In particular, Davies'
$\lambda^\circ$~\cite{davies1996temporal} has been a basis for several
multi-stage calculi with quasi-quotation. $\lambda^\circ$ did not have
operators for run and CSP; a few
studies~\cite{benaissa1999logical,MoggiTahaBenaissaSheard99ESOP} enhanced and
improved $\lambda^\circ$ towards the development of a type-safe multi-stage
calculus with quasi-quotation, run, and CSP, which were proposed by Taha and
Sheard as constructs for multi-stage programming~\cite{MetaML}. 
Finally, Taha and Nielsen invented the concept of environment
classifiers~\cite{taha2003environment} and developed a typed calculus
$\lambda^\alpha$, which was equipped with all the features above in a type
sound manner and formed a basis of earlier versions of MetaOCaml. Different
approaches to type-safe multi-stage programming with slightly different
constructs for composing and running code values have been studied by Kim,
Yi, and Calcagno~\cite{DBLP:conf/popl/KimYC06} and Nanevski and
Pfenning~\cite{DBLP:journals/jfp/NanevskiP05}.

@@

% Although there are studies on combinations of multi-stage programming and
% other programming features such as mutable cells~\cite{kiselyov2016refined}
% and control operators~\cite{KameyamaKiselyovShan09PEPM,oishi2017staging}, a
% combination with dependent types has been little studied.

%%%%%% dependent types for (compile-time) code generation

The use of dependent types for code generation is argued by
Chlipala~\cite{chlipala2010ur} and Ebner et
al.~\cite{DBLP:journals/pacmpl/EbnerURAM17}.  They use inductive types
to guarantee well-formedness of generated code.  Aside from the lack
of quasi-quotation, their systems are for heterogeneous
meta-programming and compile-time code generation and they do not
support features for run-time code generation such as run and CSP, as \LMD{} does.

%%%% index/dependent types for MSP

We discuss earlier attempts at incorporating dependent types into
multi-stage programming.  Pasalic and Taha~\cite{pasalic2002tagless}
designed \(\lambda_{H\circ}\) by introducing the concept of stage into
an existing dependent type system
\(\lambda_H\)~\cite{zhong2002certified}.  However,
\(\lambda_{H\circ}\) is equipped with neither run nor CSP.  Forgarty,
Pasalic, Siek and Taha~\cite{fogarty2007concoqtion} extended the type
system of MetaOCaml with indexed types.  With this extension, types
can be indexed with a Coq term.  Chen and Xi~\cite{chen2003meta}
introduce code types augmented with information on types of free
variables in the code in order to prevent code with free variables
from being evaluated.  These systems separate the language of type
indices from the term language.  As a result, they do not enjoy
full-spectrum dependent types but are technically simpler because
there is no need to take stage of types into account.  Brady and
Hammond~\cite{brady2006dependently} have discussed a combination of
(full-spectrum) dependently typed programming with staging in the
style of MetaOCaml to implement a staged interpreter, which is
statically guaranteed to generate well-typed code.  However, they
focused on concrete programming examples and there is no theoretical
investigation of the programming language they used.

Berger and Tratt~\cite{martin2015HGRTMP} gave program logic for
\(\text{Mini-ML}^\square_e\)~\cite{DaviesPfenning01JACM}, which would
allow fine-grained reasoning about the behavior of code generators.
However, it cannot manipulate open code which ours can deal with.
